### 2011-10-19 doc/manuscript.tex
Ilan's Results section edits replaced mine. Mine is after the document at the end.

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in}
\setlength{\parskip}{\bigskipamount}
\setlength{\parindent}{0pt}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[authoryear]{natbib}
\usepackage{longtable}
\doublespacing

\newcommand{\chck}{\textbf{[CHECK]}}
\newcommand{\addfig}{\textbf{[ADD FIGURE]}}
\newcommand{\tba}{\textbf{[TO BE ADDED]}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newcommand{\lyxaddress}[1]{
\par {\raggedright #1
\vspace{1.4em}
\noindent\par}
}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

\usepackage{babel}

\begin{document}
\renewcommand\refname{Literature Cited}

\raggedright

\subsubsection*{Title: }

Inferring homologous recombination and horizontal gene transfer 
using whole genomes of \textit{Streptococcus pyogenes} 
and\textit{ S.\ dysgalactiae} subspecies \textit{equisimilis}

\subsubsection*{Authors:}

Sang Chul Choi$^{1}$, Matthew D. Rasmussen$^{1}$, Melissa Jane Hubisz$^{1}$, Ilan 
Gronau$^{1}$,
Michael J. Stanhope$^{2}$, and Adam Siepel$^{1}$

\subsubsection*{Affiliations:}

\lyxaddress{$^{1}$Department of Biological Statistics and Computational Biology,
Cornell University, Ithaca, NY 14853, USA.}

\lyxaddress{$^{2}$Department of Population Medicine and Diagnostic Sciences,
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853,
USA.}

\subsubsection*{Author for Correspondence:}

Adam Siepel, Department of Biological Statistics and Computational
Biology, Cornell University, Ithaca, NY 14853, USA, Phone: 607-254-1157,
Fax: 607-255-4698, Email: acs4@cornell.edu\clearpage{}

\subsubsection*{Abbreviations:}

group A streptococci (GAS), group C streptococci (GCS), group G streptococci
(GGS), \textit{Streptococcus pyogenes} (SPY), \textit{S.\ dysgalactiae}
subspecies \textit{dysgalactiae} (SDD), \textit{S.\ dysgalactiae} subspecies
\textit{equisimilis} (SDE)

\subsubsection*{Keywords:}

bacterial recombination, gene duplication and transfer, ClonalOrigin,
ClonalFrame, Mowgli

\subsubsection*{Version: 0.68 (Oct 15, 2011) \clearpage{}}

\section*{Abstract}

Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have originated
in those of SPY, or vice versa.  Our genome-wide survey of gene transfer
indicated that gene transfer via homologous recombination occurred more often
for the net direction from SPY to SDE.  We found evidence that virulence genes
transferred more often via non-recombining fashion without replacing parts of
destination genomes than did non-virulence genes.  A genome-wide resource of
recombining and non-recombining gene transfer for the study of the SPY and SDE
genomes is provided so that research community in the pathogenic SPY and SDE
bacteria could take advantage of it.

\section{Introduction}

Bacteria infect humans with a variety of diseases: e.g., food poisoning
(\textit{Eschericha Coli}), strep throat (\textit{Streptococcus pyogenes}),
tuberculosis (\textit{Mycobacterium tuberculosis}), cholera (\textit{Vibrio
cholera}), anthrax (\textit{Bacillus anthracis}), pneumonia
(\textit{Chlamydophila pneumoniae}), and influenza (\textit{Haemophilus
influenzae}), to name a few.  \textit{Streptococcus pyogenes} (SPY) belonging to
group A streptococci (GAS) primarily infect humans in the throat and skin
\citep{Broyles2009}. It has high prevalence throughout the world causing
diseases ranging from mild illnesses such as pharyngitis to severe invasive
streptococcal diseases including necrotizing fasciitis and toxic shock syndrome
\citep{Cunningham2000a}.  Its close relative, \textit{Streptococcus
dysgalactiae} subspecies \textit{equisimilis} (SDE) belonging to group C or G
streptococci (GCS/GGS), used to be generally regarded as a veterinary pathogen
\citep{Vandamme1996}. However, it has been reported to cause serious and
life-threatening streptococcal diseases to humans traditionally associated with
GAS \citep[e.g.,][]{Brandt2009}.  Because of the overlap in disease profile and
ecological niche of tissue sites betwen SPY and SDE, these bacteria may as
well interact with each other by exchanging parts of their genomes.

Therefore, the research topic on gene exchange between SPY and SDE has been
performed in various ways. \citet{Sachse2002} showed gene transfer from SDE to
SPY for the downstream of the genomic region of streptococcal pyogenic
exotoxins. \citet{Davies2007} investigated phage 3396 from SDE as a vehicle of
gene transfer from SDE to SPY. \citet{Davies2009} demonstrated that an
integrative conjugative element of SDE allowed gene transfer from SDE to other
streptococci.  \citet{Ahmad2009} did not conclusively deduce the net
directionality of housekeeping gene transfer between SPY and SDE.  Notably,
\citet{Kalia2001} attempted to show that the net direction of gene transfer from
SPY to SDE was more dominant than the other direction.  Because these researches
on the controversial issue, net directionality of gene transfer between SPY and
SDE, have been limited to relatively small regions in the genomes primarily on
representative loci from the Multi Locus Sequence Typing (MLST) database, it
remains to be elusive to determine the genome-wide degrees to which genomic
segments flowed between the two species, and net directions of thereof.

As sequencing costs plummet \citep{Mardis2011}, bacterial genome sequencing
produces individual genomes of high quality even within a bacterial species
\citep{Tettelin2009a}.  Studies on horizontal gene transfer in prokaryotes have
used genes in very large sets of species \citep{Koonin2001} from a great deal of
genomes available.  Although this type of study gave us insights into deep
evolutionary histories of prokaryotes, with multiple individual genomes per
species we now can go beyond such rough broad study on gene transfer towards
fine-scaled history of gene transfer between very closely related prokaryotes.
For instance, \citet{Liu2009} studied gene transfer events in
\textit{Lactobacillus bulgaricus} and \textit{Streptococcus thermophilus} using
the compositional approach of \citet{Karlin2001}. \citet{Luo2011} investigated
gene transfer events in \textit{Eschericha Coli} using the embedded quartet
decomposition due to \citet{Zhaxybayeva2006}.  \citet{Hamady2006} combined
phylogenetic and compositional approaches to detecting horizontal gene transfer.
\citet{Caro-Quintero2011} studied a number of \textit{Shewanella baltica}
genomes to identify the genetic elements that enabled the species to adapt to
redox gradients.  Although several complete genomes from SPY have been
available, \citet{Suzuki2011} just recently added one more complete genome of
SDE to the set of a previously unique SDE genome \citep{Shimomura2011}. These
multiple SDE genomes prompted us to study genome-wide gene flows between SPY and
SDE that afflicted human beings around the globe.

@@

Homologous recombining gene transfer could be studied more readily using the
core genomes of individual bacteria while horizontal gene transfer might be more
easily detected using the pan genomes. Recently, a method of inferring
homologous recombination with whole bacterial genomes was developed
\citep{Didelot2010}. The model of homologous recombination allowed them to study
flows of genetic segments between bacteria individuals of \textit{Bacillus
cereus} species in the core genomes.  Other evolutionary processes acting on
bacterial genomes include genome rearragement, translocation, inversion,
duplication, gain, loss, etc.  These processes did not readily lend themselves
to statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene duplication, loss, and
transfer events was developed by \citet{Doyon2011}.  We noted that the first
approach could be used to infer transfer of genomic segments due to homologous
recombination, and the second could be useful in studying horizontal gene
transfer.

In short, the complete genomes of the two species of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. Virulence genes were found to be more associated
non-recombining horizontal gene transfer than non-virulence genes.  We compared
both of the results in recombining gene transfer events to show that the
parsimony-based method inferred recombining gene transfer events in biological
functional categories similar to those inferred with the model-based approach.
This was the first attempt as far as we know of systematically analyzing SPY and
SDE in genome-wide scale to detect gene transfer events not only in the core
genome but also in the pan genome, which made the analysis complete in terms of
the amount of used genome data. Our analysis of core genomes also showed for the
first time that the net directionality of recombining gene transfer from SPY to
SDE was stronger than the reverse direction.

\section{Results}

\subsection{Recombination drived more substitutions than did point mutation}

Five genomes except \textit{S.\ equi} ssp.\textit{\ equi} strain 4047 were
chosen (Table S\ref{tab:genome}) to study recombining gene transfer using the
model-based approach.  A genome alignment of the five genomes showed two
distinct genome structures of SPY and SDE (Figure S\ref{fig:mauve}). The SDE and
SDD genomes were larger in length than those of SPY. Although the graphical view
of the genome alignment did not clearly show whether SDD was different from
either of SDE1 and SDE2 in the genome organization, a spcies tree inferred from
the genome alignment showed that the external SDD branch was attached to the
internal clade branch of SDE1 and SDE2 (Figure S\ref{fig:tree5}). The
distribution of the alignment block lengths was skewed to the right with mean of
4188 base pairs long (Figure S\ref{fig:blocksize}), which supported that the
model assumption of geometric block length distribution might be plausible in
\texttt{ClonalOrigin} method (see section \textbf{Materials and Methods} and
\citealt{Didelot2010}).  The relative strength ($r/m$) of recombination compared
to mutations indicated that recombination-driven substitutions affected more
sites of the core genomes than did mutation-driven substitutions (Table
S\ref{tab:clonalframe}). This could imply that in the \textit{Streptococcus}
species bacterial recombination process appeared to accelerate molecular
evolution of more sites than did point mutations process.

\subsection{Recombination hotspots were observed in genomes}

\subsection{SPY and SDE were more closely related in about tenth of the core genome}

\subsection{Unbalaned recombining gene flow from SPY to SDE in genome-wide scale}

We tried in vain to quantify differential horizontal gene transfer between SPY
and SDE with the parsimony-based approach (Figure \ref{fig:hgt-heatmap}).  Among
the clade-crossing transfers, we did not see any significant enrichment for a
specific direction, as opposed to the result for recombining transfers from the
model-based analysis.  SDE1 and SDE2 shared the most events (26 and 34) compared
to other sample pairs.  SDD appeared to acquire many genes from the ancestral
SDE branch.  Transfer events between the SPY and SDE clades were inferred in
smaller numbers (<6) compared to those within each clade. We were puzzled by
this finding because virulence genes were found to be associated with horizontal
gene transfer, and yet we could not show net directionality of horizontal gene
transfer between SPY and SDE.  However, we inferred relatively more recombining
gene transfer events from the SPY branch to SDE branch even in the
parsimony-based approach (Figure S\ref{fig:mowgli-recomb-heatmap}).

\subsection{Reduction of genes and divergence of SPY and SDE species}

@@

\subsection{Virulence genes via recombining gene transfer}

>>>>

% Four potential reviewers whose e-mail addresses you can provide
% Claus O. Wilke
% Debra Bessen
% Bruce R. Levin
% Michael Wessels

% Template for PLoS
% Version 1.0 January 2009
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

% FIXME: Remove this package when submitting the manuscript.
\usepackage{url}

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}

% FIXME: Remove the followings if you have to when submitting the manuscript.
\let\citep\cite
\let\citet\cite
\providecommand{\tabularnewline}{\\}
\newcommand{\lyxdot}{.}
\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}

\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Inferring homologous recombination and horizontal gene transfer 
using whole genomes of \textit{Streptococcus pyogenes} 
and\textit{ S.\ dysgalactiae} ssp.\textit{\ equisimilis}}
}
% Short Title (running head): Gene transfer using Streptococcus genomes
% Insert Author names, affiliations and corresponding author email.
\\
Sang Chul Choi$^{1}$, Matthew D. Rasmussen$^{1}$, 
Melissa Jane Hubisz$^{1}$, Ilan Gronau$^{1}$,
Michael J. Stanhope$^{2}$, and Adam Siepel$^{1,\ast}$
\\
\bf{1} 
Department of Biological Statistics and Computational Biology,
Cornell University, Ithaca, NY 14853, USA
\\
\bf{2}
Department of Population Medicine and Diagnostic Sciences,
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\
$\ast$ E-mail: acs4@cornell.edu, Phone: 607-254-1157, Fax: 607-255-4698
\end{flushleft}

\section*{Things To Do for submitting this to PLoS Pathogens}
\begin{enumerate}
\item A paragraph for briefing the methods of ClonalOrigin and Mowgli before
section \textbf{Results}.
\item We can add at least 50 words in section \textbf{Abstract}.
\item Figures, tables, and supporting information need to be ready.
\item Publish source codes at public domain.
\end{enumerate}

\section*{A list our major and minor findings or observations}
\begin{enumerate}
\item We showed that recombining gene transfer from SPY to SDE was strong
enough.
\item A few evidences for the net directionality from SPY to SDE:
time-constrained gene transfer between SPY to SDE also showed the directionality
from SPY to SDE.
\item We showed that virulence genes were associated with horizontal gene
transfer. 
\item We found that recombining gene tranfer processes are involved in
membrane-bound protein encoding genes and aminoacyl tRNA synthetases.
\item We found that horizontal gene transfer processes are involved in
transposases.
\item We provided evidence that recombination process affected more sites along
the core genome of SPY and SDE than mutation process.
\item We found a recombination hotspot at 200 kb from the origin of replication
along the genome of SDE1.
\item We showed that species-tree-topology-changing recombinant edges were more
reliably recovered than species-tree-topology-not-changing edges.
\item We showed that recombination intensity might be useful in finding genes
under high recombining gene transfer.
\item We found that SPY and SDE are more closely related than SDD and SDE in
about 10\% of the core genome.
\item Recombination intensity and number of recombining gene transfer events
were positively correlated.
\item We found reduction of numbers of genes after divergence of SDE and SPY.
\item We confirmed the species tree that relates the five genomes.
\item We listed several virulence genes associated with recombining and
horizontal gene transfer.
\end{enumerate}

\section*{Suggestions about the manuscript}
\begin{enumerate}
\item Ilan thought that a species tree must be shown at the outset of the
section \textbf{Results}.
\item Ilan suggested a different order of four parts in \textbf{Results}:
preliminary, funtional category and virulence gene analyses, gene gain and loss,
and directionality.
\end{enumerate}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have originated
in those of SPY, or vice versa.  Two mechanisms of gene transfer in bacteria
were known; recipient bacteria of transferred genes can substitute transferred
genes for homologous parts of their genomes (Recombining Gene Transfer), or
integrate transferred genes into their genomes without replacement (Horizontal
Gene Transfer).  We discriminated the two processes by using a model-based
approach for studying recombining gene transfer, and using a parsimony-based
approach for both transfer processes.  Our genome-wide survey of gene transfer
indicated that gene transfer via homologous recombination occurred more often
for the net direction from SPY to SDE.  We also found evidence that virulence
genes transferred more often via non-recombining fashion without replacing parts
of destination genomes than did non-virulence genes.  
A genome-wide resource of
recombining and non-recombining gene transfer for the study of the SPY and SDE
genomes is provided so that research community in the pathogenic SPY and SDE
bacteria could take advantage of it.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

\textit{Streptococcus pyogenes} (SPY) has afflicted people around the globe
causing many diseases. Its closely related species, \textit{S.\ dysgalactiae}
subspecies \textit{equisimilis} (SDE), used to be known as veterinary pathogens
have been reported to share disease profiles with SPY.  While gene transfers
between the two species were shown in small portions of genomic loci, no study
on the gene transfer at genome-scale has been performed because of lack of
methodologies implemented or genomic data available. Recent advances in both
method development and genomic data allowed to study gene transfer between the
two species in genome scale.  Recipient bacteria of transferred genes can
substitute transferred genes for homologous parts of their genomes, or integrate
transferred genes into their genomes without replacement.  The former is called
recombining gene transfer, and the latter horizontal gene transfer. By
discriminating the two processes, We found that virulence genes were more
associated with horizontal gene transfer than recombining gene transfer.  We
also found a signficant signal of recombining gene transfer for the net
direction from SPY to SDE in the core genome.  Our study provides new insights
into gene transfer between SPY snd SDE in genome scale.

\section*{Introduction}

Bacteria infect humans with a variety of diseases: e.g., food poisoning
(\textit{Eschericha Coli}), strep throat (\textit{Streptococcus pyogenes}),
tuberculosis (\textit{Mycobacterium tuberculosis}), cholera (\textit{Vibrio
cholera}), anthrax (\textit{Bacillus anthracis}), pneumonia
(\textit{Chlamydophila pneumoniae}), and influenza (\textit{Haemophilus
influenzae}), to name a few.  \textit{Streptococcus pyogenes} (SPY) belonging to
group A streptococci (GAS) primarily infect humans in the throat and skin
\citep{Broyles2009}. It has high prevalence throughout the world causing
diseases ranging from mild illnesses such as pharyngitis to severe invasive
streptococcal diseases including necrotizing fasciitis and toxic shock syndrome
\citep{Cunningham2000a}.  Its close relative, \textit{Streptococcus
dysgalactiae} subspecies \textit{equisimilis} (SDE) belonging to group C or G
streptococci (GCS/GGS), used to be generally regarded as a veterinary pathogen
\citep{Vandamme1996}. However, it has been reported to cause serious and
life-threatening streptococcal diseases to humans traditionally associated with
GAS \citep{Brandt2009}.  Because of the overlap in disease profile and
ecological niche of tissue sites betwen SPY and SDE, these bacteria may as
well interact with each other by exchanging parts of their genomes.

Therefore, the research topic on gene exchange between SPY and SDE has been
performed in various ways. Sachse et al.\ \citet{Sachse2002} showed gene transfer from SDE to
SPY for the downstream of the genomic region of streptococcal pyogenic
exotoxins. Davies et al.\ \citet{Davies2007} investigated phage 3396 from SDE as a vehicle of
gene transfer from SDE to SPY. They also \citet{Davies2009} demonstrated that an
integrative conjugative element of SDE allowed gene transfer from SDE to other
streptococci. Ahmad et al.\ \citet{Ahmad2009} did not conclusively deduce the net
directionality of housekeeping gene transfer between SPY and SDE.  Notably,
Kalia et al.\ \citet{Kalia2001} attempted to show that the net direction of gene transfer from
SPY to SDE was more dominant than the other direction.  Because these researches
on the controversial issue, net directionality of gene transfer between SPY and
SDE, have been limited to relatively small regions in the genomes primarily on
representative loci from the Multi Locus Sequence Typing (MLST) database, it
remains to be elusive to determine the genome-wide degrees to which genomic
segments flowed between the two species, and net directions of thereof.

As sequencing costs plummet \citep{Mardis2011}, bacterial genome sequencing
produces individual genomes of high quality even within a bacterial species
\citep{Tettelin2009a}.  Studies on horizontal gene transfer in prokaryotes have
used genes in very large sets of species \citep{Koonin2001} from a great deal of
genomes available.  Although this type of study gave us insights into deep
evolutionary histories of prokaryotes, with multiple individual genomes per
species we now can go beyond such rough broad study on gene transfer towards
fine-scaled history of gene transfer between very closely related prokaryotes.
For instance, Liu et al.\ \citet{Liu2009} studied gene transfer events in
\textit{Lactobacillus bulgaricus} and \textit{Streptococcus thermophilus} using
the compositional approach of Karlin \citet{Karlin2001}. Luo et al.\ \citet{Luo2011} investigated
gene transfer events in \textit{Eschericha Coli} using the embedded quartet
decomposition due to Zhaxybayeva et al.\ \citet{Zhaxybayeva2006}. Hamady et al.\ \citet{Hamady2006} combined
phylogenetic and compositional approaches to detecting horizontal gene transfer.
Caro-Quintero et al.\ \citet{Caro-Quintero2011} studied a number of \textit{Shewanella baltica}
genomes to identify the genetic elements that enabled the species to adapt to
redox gradients.  Although several complete genomes from SPY have been
available, Suzuki et al.\ \citet{Suzuki2011} just recently added one more complete genome of
SDE to the set of a previously unique SDE genome \citep{Shimomura2011}. These
multiple SDE genomes prompted us to study genome-wide gene flows between SPY and
SDE that afflicted human beings around the globe.

@@

Recently, a method of inferring
homologous recombination with whole bacterial genomes was developed
\citep{Didelot2010}. The model of homologous recombination allowed them to study
flows of genetic segments between bacteria individuals of \textit{Bacillus
cereus} species in the core genomes.   
% Descriptions of ClonalOrigin model.
Other evolutionary processes acting on
bacterial genomes include genome rearragement, translocation, inversion,
duplication, gain, loss, etc.  These processes did not readily lend themselves
to statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene duplication, loss, and
transfer events was developed by Doyon et al.\ \citet{Doyon2011}.  
% Descriptions of Mowgli analsysis.
We noted that the first
approach could be used to infer transfer of genomic segments due to homologous
recombination, and the second could be useful in studying horizontal gene
transfer.
% Homologous recombining gene transfer could be studied more readily using the
% core genomes of individual bacteria while horizontal gene transfer might be more
% easily detected using the pan genomes.

In short, the complete genomes of the two species of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. Virulence genes were found to be more associated
non-recombining horizontal gene transfer than non-virulence genes.  We compared
both of the results in recombining gene transfer events to show that the
parsimony-based method inferred recombining gene transfer events in biological
functional categories similar to those inferred with the model-based approach.
This was the first attempt as far as we know of systematically analyzing SPY and
SDE in genome-wide scale to detect gene transfer events not only in the core
genome but also in the pan genome, which made the analysis complete in terms of
the amount of used genome data. Our analysis of the core genomes also showed for the
first time that the net directionality of recombining gene transfer from SPY to
SDE was stronger at the genome scale.

% Results and Discussion can be combined.
\section*{Results}

\subsection*{Distinguishing two gene transfer processes shows different functional
categories associations}

Bacterial gene transfer processes as an evolutionary force would affect more
frequently in some biological functions than in others.  In our attempt of
finding which biological functions were associated with recombining or
horizontal gene transfers, we devised a measure of recombination intensity for
genes by using frequency of recombinant edges that affect those genes, and used
frequency of horizontal gene transfer events (see section \textbf{Materials and
Methods}).  Table \ref{tab:functional} shows a list of function categories
that were found to be associated with recombining gene transfers. We noted three
biological functions in the list: membrane-bound molecules, aminoacyl tRNA
synthetases, and ribosomes.

Firstly, we have sugar phosphotransferases ranked
top in the list; phosphrylation could be important in signal transduction to
recognize environmental changes.  Transport functional categories of
carbohydrate transport and sodium ion transport would belong to genes encoding
membrane-bound molecules.  Also highly ranked were fatty acid and lipid
biosyntheses that must be involved in producing membrane-bound molecules. 
Because bacteria as a unicellular organism should be able to respond to changes
to environment, some evolutionary pressures must have been acting on genes
encoding membrane-bound molecules. Therefoure, recombining gene transfer process
appeared to be more associated with biological functions that had been critical
of sensing queues of bacterial environmental changes.

Secondly, we found several aminoacyl tRNA synthetases (GO Term translation
GO:0006412) in association with recombining gene transfer.  In the
parsimony-based approach the term translation was also top ranked (Table
S\ref{tab:go-events-recombining}).  Because they had been a focus of gene
transfer in bacterial species \citep{Woese2000}, we could confirmed that the
recombination intensity and requency of recombining gene transfer events would
be useful in finding genes associated with recombining gene transfer.

Thirdly, functional categories concerning ribosomes were unexpectedly
found to be associated with recombining gene transfer.  Because genes related
with ribosomes were considered to be housekeeping genes, gene transfer in which
would be rare, it was unexpected that ribosome related genes were found to be
associated with recombining gene transfer.  Because it was a bizarre result, we
investigated the association by considering different recombinant edge types. We
noted that two different classes of recombinant edge types: one that changes
species tree topologies, and another that does not
(see section \textbf{Materials and Methods}).
Because recombinant edges
that did not change species tree topology might be difficult to infer reliably,
considering those recombinant edges in computing recombination intensity might
lead to dubious results.  Ribosome related functional categories were found to
be top ranked if we took those recombinant edges that did not change species
tree topology (Tables S\ref{tab:functional-notopology} and
S\ref{tab:functional-all}).  In Table \ref{tab:functional} ribosome related
functional categories were not top ranked.  Therefore, high recombination
intensity in ribosome related genes were likely to be attributed to
topology-not-changing recombinant edges. We would be cautious of recombining
transfer of genes related with ribosome related functions.

Table \ref{tab:go-events} shows a list of functional categories associated with
horizontal gene transfer, which were different from those from recombining gene
transfer analysis.  Ranked top in the list was transposases, which were known to
help foreign genes be excised from a donor genome and integrated into a
recipient genome. The biological role of transposases was concordant with
horizontal gene transfer process because the process involved in transposases
would allow bacteria to acquire foreign genes without the requirement of
homologous regions in the bacteria.  
These differences in the profile of functional categories association of the two
gene transfer processes were because of the approach of dicriminating the two
gene transfer processes.

\subsection*{Recombining gene transfer from SPY to SDE at genome-wide scale}

We show a net directionality of recombining gene transfers from SPY to SDE.

Describe our arguments of the net directionality. 
Number of recombinant edges.
Confidence about the numbers of recombinant edges.
Not all of genes experience recombining gene transfers from SPY to SDE.
Gene transfer counts from the parsimony-based approach; the power of the
parsimony-based approach may be limited.
Is this number of recombinant edges good enough for the argument?

Reflection of the previous analyses, 
heat maps of real data analysis, 
and supporting simulations.

\subsection{Bayesian analysis of recombining gene transfer}

{\bf IG: REMOVED A LOT OF DETAIL (MAINLY IN FIRST PART). PLEASE CHECK.}

We followed the pipeline and methods suggested in \cite{Didelot2010} to pursue a model based study of recombining gene transfer. A preliminary step in this analysis involved obtaining an estimate of the species tree with branch lengths (via ClonalFrame), and estimates of the three population parameters: mutation rate, recombination rate, and average recombinant tract length (via a first phase of ClonalOrigin). Due to divergence of the SEE sample from the other five, we could not obtain suficiently long reliable multiple sequence alginments with it, and had to exclude it from the bayesian analysis. The spcies tree inferred by ClonalFrame over the five species has the expected topology (Fig. {\bf [ADD FIGURE 1]} (see Figure S\ref{fig:tree5} for branch length estimates). ClonalFrame also produces estimates for the relative strength of recombination compared
to mutations ($r/m$; Table S\ref{tab:clonalframe}).  The estimate of $r/m=4.66$ provides an initial indication to the important part recombination took in the evolutionary process of shaping these genmes: four times more sites have been affected by recombination-driven substitutions than have been affected by mutation-driven substitutions. Estimates of population mutation rate ($\theta$), recombination rate ($\rho$), and average tract length ($\delta$) were obtained for the different alignment blocks analyzed (Figure S\ref{fig:scatter3}). Global estimates of these three population parameters were obtained by a length-weighted average (Materials and Methods). The local estimates typically seem to be moderately fluctuating around the global average, with the exception of what apears to be a recombination hotspot at a region 200 kilobase pairs from the origin of replication of SDE1.

The species tree and estimated branch lengths, together with the global estimates of the three population parameters are used in the second phase of ClonalOrigin, where recombinant trees are sampled from the posterior distriution. One way to summarize the posterior sample of recombinant trees provided by ClonalOrigin is to observe the topologies of the local genealogies implied by it. Out of the 105 possible rooted labeled topologies spanning five taxa, the dominant topology (sampled 66.7\% of the time) was that of the species tree (Materials and Methods; Table S\ref{tab:Gene-tree-topologies}). The second most frequently sampled topology (sampled 9.3\% of the time) was the one in which SPY and SDE are sister clades (with SDD as an outgroup). Other topologies were sampled with frequency less than 4\%. This gives initial indication to the extent of recombining gene transfer  between SPY and SDE.

To gain a more fine-scale view of the main types of recombination scenarios sampled by ClonalOrigin, we looked at the recombinant edges present in the sampled recombinant trees. Following the approach of \cite{Didelot2010}, we classified these edges by types defined by distinct pairs of donor and recepient branches of the species tree (Fig. [ADD FIG 1]). For each edge type, we recorded the average number of edges of that type present in the sampled recombinant tree, and computed the ratio of this observed count and the prior expected count as implied by the species tree (considering branch lengths) and the three population parameters estimated in the preliminary round of the analysis. The base 2 logarithms of these ratios are depicted in the heatmap of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap}).

Among the possible recombinant edge types, we distinguished between ones which preserve the species tree topology, and ones which alter it. Our main focus was on the topology-altering types, both because our main objective was to study cross-clade gene transfer, and because our simulation study indicates that the observed counts of topology-preserving recombinant edges are somewhat less reliable {\bf[IG: POINT TO RELEVANT SECTION]}. The heatmap allows us to find recombination scenarios that deviate from the expectation dictated by the probabilistic model of ClonalOrigin. {\bf [IG: CHECK THIS FOLLOWING LINE OF ARGUMENT]} We identify two main potential causes for such a deviation: (1) reduced or increased opportunities for species to interact and exchange genetic material; (2) selection acting on newly transferred genes.

We observe a reduced count of recombinant edges sampled between SDD and the three branches of the SPY clade ($\sim$25\% of the expected count). This is consistent with the fact that these species have reduced opportunities to interact, SPY being a human pathogen and SDD being strictly veterinary. On the contrary, the number of sampled recombinant edges between SDD and branches in the SDE clade seems to be consistent with the prior, with a slight excess for edges from SDE1 and SDE2 to SDD {\bf [IG: IS THIS WORTH DWELLNG ON?]}. The greatest deviation of observed counts from expected ones was observed for recombinant edges between the SPY and SDE clades. This is consistent with the fact that these two species share an ecological niche (SDE being an opportunistic human pathogen). Interestingly, there appears to be much higher enrichment (up to four times more than expected) for recombinant edges originating in the SPY clade and ending up in the SDE clade. In a follow-up simulation study we show that this deviation from the expected count is indeed highly significant, and so is the observed asymmetry in pattern of recombination {\bf[IG: POINT TO RELEVANT SECTION]}. This bias in directionality of recombination pattern between SPY and SDE is especially intriguing in light of the hypothesis that SDE acquired human pathogenicity via gene transfer from SPY.

\subsection{Parsimony-based reconciliation of gene trees with the species tree}

{\bf[IG: DID NOT TOUCH THE FIRST 3 PARAGRAPHS. REQUIRES ``CONDENSING".]}

{\bf[IG: START WITH SOME GENERAL STATISTICS - NUMBER OF GENES, ETC?]}

@@

{\bf [IG: MODIFIED THIS PARAGRAPH]}

As in the Bayesian analysis of ClonalOrigin, we attempted to see what types of transfer events (horizontal and recombining) were most common in our inference. We classified events into types defined by donor-recepient branches in the species tree. We recorded the number of inferred events of each type in Figures \ref{fig:hgt-heatmap} \&S\ref{fig:mowgli-recomb-heatmap} {\bf [IG: CONSIDER MOVING BOTH INTO SUPPLEMENT]}. Unlike in the Bayesian analysis, we cannot normalize these counts by their expectation, since we do not have any model-based expectation for them. This restrict our ability to interprest these counts. We did notice, however, a very high number of inferred recombining gene transfer events from the ancestral SPY branch to the ancestral SDE branch, consistent with our observations from the Bayesian analysis. With horizontal gene transfer, we do not see a similar pattern, but we do observe a high number of horizontal transfers from the ancesral SDE branch to SDD. {\bf [IG: IS THERE ANYTHING ELSE WE CAN SAY ABOUT THIS? MAYBE SHOW A SCATTER PLOT OF THE TWO TYPES OF EVENTS? ]}

\subsection{Gene categories associated with recombining and horizontal gene transfer}

%\subsection{Association of virulence genes and horizontal gene transfer}
